### 2013-06-02 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
starting work on spn solvers comparison section

\subsection{Monte Carlo Estimator Comparison}
\label{subsec:spn_estimator_comparison}
With convergence obtained for the fuel assembly problem, the advanced
preconditioners will next be studied with both the collision and
expected value estimators using the best preconditioning and
relaxation parameter combinations found by the previous
analysis. Additionally, both the Richardson-based MCSA iteration and
the MINRES-based MCSA iteration given by Eq~(\ref{eq:mcsa_minres})
will be used. Although the expected value estimator has achieved the
best iterative performance in the previous comparison and analysis for
the simple Poisson problem in
\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective due
to the sparsity lost by explicit preconditioning. The deterministic
component of the expected value estimator that couples the current
state to other local states through the iteration matrix stencil
requires application of the estimator to potentially several orders of
magnitude more states at every transition event.

For this study, the fuel assemby problem with a 1-group $SP_1$
discretization was solved using MCSA with the adjoint Monte Carlo
solver using both the collision and expected value estimators and both
the Richardson and MINRES fixed point iterations. Using the results
from the previous section, a Neumann relaxation parameter of 0.7 was
used with both estimators and fixed point iterations. For the
collision estimator, a Richardson relaxation parameter of 1.1 was used
with the Richardson iteration while it was found the expected value
estimator had the best performance when a a relaxation parameter of
1.0 was used instead. In addition, ILUT preconditioning with a drop
tolerance of \sn{1}{-5} and a fill level of 5 was used with a reduced
domain approximation fill level of 100 and a threshold of \sn{1}{-10}
as in the previous analysis. For each estimator, the number of
stochastic histories used to compute the MCSA correction was varied
from 25 to 50,000 with the number of MCSA iterations required to
converge to a tolerance of \sn{1}{-8} recorded for a single eigenvalue
iteration along with the CPU time required to achieve convergence.

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_time} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator with both fixed point
iterations creates a relative insensitivity to the number of histories
used to compute the correction. For the collision estimator, the
MINRES iteration permits fewer stochastic histories to be used with
each MCSA iteration while still maintaining good iterative
performance. With respect to CPU time, there is a more pronounced
effect due to the larger density of states created by the explicit
ILUT preconditioning. Using a reduced domain fill level of 100 creates
a situation where the expected value estimator is significantly more
expensive per history to compute that the collision estimator due to
the coupling of states.

@@

From both Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:spn_estimator_time} we see that both estimators have
comparable iterative and timing performance when 1,000 histories are
used at each MCSA iteration. Using 1,000 histories per iteration, the
infinity norm of the residual vector is presented at every MCSA
iteration for both estimators in
Figure~\ref{fig:spn_estimator_convergence} as an additional means of
comparison. At this number of histories, both estimators converge the
fuel assembly problem monotonically at and approximately the same
rate, giving little reason to choose one over the other. The end
result is that both estimators provide a viable method for getting
good MCSA iterative performance with reasonable timing
performance. For the collision estimator, using around 20,000
stochastic histories at every iteration gave the best iterative
performance while 500 histories per iteration gave the best iterative
performance when using the expected value estimator. For purely serial
computational performance, there is not a distinguishing feature for
the fuel assembly problem presented that would cause one to choose one
of the estimators over the other.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Comparison to Conventional Methods}
\label{sec:spn_comparison}

\subsection{MCSA Verification}
\label{subsec:spn_mcsa_verification}

\subsection{Performance Results}
\label{subsec:spn_comparison_results}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
  \end{center}
  \caption{\textbf{Average number of iterations required to converge
      each eigenvalue iteration for the fuel assembly problem as a
      function of energy groups.} \textit{All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:comparison_iterations}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per iteration in seconds for the
      fuel assembly problem as a function of energy groups.}
    \textit{All linear solver iterations over all eigenvalue
      iterations were used to compute the averge. All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:comparison_time}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per iteration in seconds for the
      fuel assembly problem as a function of energy groups with
      preconditioning time included for the MCSA methods.}
    \textit{All linear solver iterations over all eigenvalue
      iterations were used to compute the averge. All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:comparison_prec_time}
\end{figure}

>>>>

\subsection{Monte Carlo Estimator Comparison}
\label{subsec:spn_estimator_comparison}
With convergence obtained for the fuel assembly problem, the advanced
preconditioners will next be studied with both the collision and
expected value estimators using the best preconditioning and
relaxation parameter combinations found by the previous
analysis. Additionally, both the Richardson-based MCSA iteration given
by Eq~(\ref{eq:mcsa}) and the MINRES-based MCSA iteration given by
Eq~(\ref{eq:mcsa_minres}) will be used. Although the expected value
estimator achieved the best iterative performance in the previous
comparison and analysis for the simple Poisson problem in
\S~\ref{sec:parameter_estimator_analysis}, it is possible that the
collision estimator may perform better from a timing perspective due
to the sparsity lost by explicit preconditioning. The deterministic
component of the expected value estimator that couples the current
state to other local states through the iteration matrix stencil
requires application of the estimator to potentially several orders of
magnitude more states at every transition event.

For this study, the fuel assemby problem with a 1-group $SP_1$
discretization was solved using MCSA with the adjoint Monte Carlo
solver using both the collision and expected value estimators and both
the Richardson and MINRES fixed point iterations. Using the results
from the previous section, a Neumann relaxation parameter of 0.7 was
used with both estimators and fixed point iterations. For the
collision estimator, a Richardson relaxation parameter of 1.1 was used
with the Richardson iteration while it was found the expected value
estimator had the best performance when a Richardson relaxation
parameter of 1.0 was used instead. In addition, ILUT preconditioning
with a drop tolerance of \sn{1}{-5} and a fill level of 5 was used
with a reduced domain approximation fill level of 100 and a threshold
of \sn{1}{-10} as in the previous analysis. For each estimator, the
number of stochastic histories used to compute the MCSA correction was
varied from 25 to 50,000 with the number of MCSA iterations required
to converge to a tolerance of \sn{1}{-8} recorded for a single
eigenvalue iteration along with the CPU time required to achieve
convergence.

Figure~\ref{fig:spn_estimator_iters} gives the iteration count results
and Figure~\ref{fig:spn_estimator_time} gives the timing
results. Remarkably, although the fuel assembly problem is
significantly more complicated than the simple Poisson problem studied
in \S~\ref{sec:parameter_estimator_analysis}, the results here are
largely the same with Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:estimator_nh_iters} having identical qualitative
behavior. Using the expected value estimator with both fixed point
iterations creates a relative insensitivity to the number of histories
used to compute the correction. Marginally better iterative and timing
performance was observed for the expected value estimator when using
the Richardson iteration. Better timing performance is expected as
computing the extrapolation parameter in the MINRES iteration requires
several additional parallel operations. However, we expect that
because the MINRES iteration will converge faster on its own when
compared to the Richardson iteration that better iterative performance
will be observed when it is accelerated with MCSA. The fact that this
was not observed signals that an MCSA scheme leveraging the expected
value estimator is also effectively insensitive to the fixed point
iteration used. For the collision estimator, the MINRES iteration
permits fewer stochastic histories to be used with each MCSA iteration
while still maintaining good iterative performance. With respect to
CPU time, there is a more pronounced effect due to the larger density
of states created by the explicit ILUT preconditioning. Using a
reduced domain fill level of 100 creates a situation where the
expected value estimator is significantly more expensive per history
to compute that the collision estimator due to the coupling of states.

@@

From both Figure~\ref{fig:spn_estimator_iters} and
Figure~\ref{fig:spn_estimator_time} we see that both estimators and
fixed point iterations have comparable iterative and timing
performance when 1,000 histories are used at each MCSA
iteration. Using 1,000 histories per MCSA iteration, the infinity norm
of the residual vector is presented at every MCSA iteration for both
estimators and both fixed point iterations in
Figure~\ref{fig:spn_estimator_convergence} as an additional means of
comparison. At this number of histories, all MCSA combinations
converge the fuel assembly problem monotonically at and approximately
the same rate, giving little reason to choose one over the other. The
end result is that both estimators provide a viable method for getting
good MCSA iterative performance with comparable timing
performance. For the collision estimator, using around 20,000
stochastic histories at every iteration gave the best iterative
performance while 500 histories per iteration gave the best iterative
performance when using the expected value estimator. For purely serial
computational performance, there is not a distinguishing feature for
the fuel assembly problem presented that would cause one to choose one
of the estimators and fixed point iterations over the other as they
all have similar minimum CPU times over the range of values tested.

@@

%%---------------------------------------------------------------------------%%
\section{MCSA Comparison to Conventional Methods}
\label{sec:spn_comparison}
Through the numerical studies presented up to this point in this
chapter, we have developed a set of preconditioning techniques that
permit MCSA to be used with the $SP_N$ form of the neutron transport
equation and applied to a difficult fuel assembly criticality
problem. Along with these preconditioning techniques, a set a varying
MCSA parameters were analyzed to find those that yielded the best
performance for this particular problem. In addition, several steps
were taken to mitigate the dense composite linear operators that arise
when performing explicit MCSA preconditioning in order to achieve
convergence. Using these results, we compare MCSA directly to
conventional linear solvers that would be typically used to solve the
$SP_N$ form of the transport equation as presented here. In
particular, we will use a pair of production Krylov solvers with which
we will compare direct numerical results in order to verify MCSA
solutions and iterative and CPU timing performance in order to gauge
feasibility and identify areas that require further research.

\subsection{MCSA Verification}
\label{subsec:spn_mcsa_verification}
In order to verify the correctness of the MCSA method and
implementation used for this work, MCSA solutions to the fuel assembly
problem with 1, 2, and 4 energy groups using $SP_1$ discretization
will be directly compared over all group and cell fluxes computed by
converging the criticality problem using both BiCGStab and GMRES
production implementations from the Trilinos scientific computing
library Aztec \citep{heroux_overview_2005}. In addition to comparing
the cell and group fluxes, the number of eigenvalue iterations
required to converge the problem along with the eigenvalue itself will
be compared as a means of verification. For the verification, all
solvers will be preconditioned with ILUT using a drop tolerance of
\sn{1}{-5} and a fill level of 5. For GMRES, no restrictions were
placed on the size of the subspace and therefore no restarts
occurred. When the collision estimator was used with MCSA, 20,000
stochastic histories were used to compute the correction for every
energy group in the problem (40,000 and 80,000 histories total for the
2 and 4 group calculations respectively) and 500 stochastic histories
used for each energy group when the expected value estimator was used
(1,000 and 1,500 histories total for the 2 and 4 group
calculations). The reduced domain approximation was also applied in
conjuction with the ILUT preconditioning using a fill level of 100 and
a threshold value of \sn{1}{-10} to reduce the density of states in
the Monte Carlo problem. For relaxation parameters, all MCSA
compuations used a Neumann relaxation parameter of 0.7 while a
Richardson relaxation of 1.1 was used with the collision estimator and
1.0 with the expected value estimator as determined by the previous
analysis. Table~\ref{tab:spn_solver_defs} gives definitions for the
solvers used to generate the results in the remainder of this section.
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{l}{\textbf{Name}} & 
      \multicolumn{1}{l}{\textbf{Definition}} \\
      BiCGStab-ILUT & ILUT preconditioned BiCGStab \\
      GMRES-ILUT & ILUT preconditioned GMRES \\
      MCSA-ILUT-R-C & ILUT preconditioned MCSA with Richardson \\ 
                    & fixed point iteration and collision estimator \\
      MCSA-ILUT-MR-C & ILUT preconditioned MCSA with MINRES \\
                     & fixed point iteration and collision estimator \\
      MCSA-ILUT-R-EV & ILUT preconditioned MCSA with Richardson \\
                     & fixed point iteration and expected value estimator \\
      MCSA-ILUT-MR-EV & ILUT preconditioned MCSA with MINRES \\
                      & fixed point iteration and expected value estimator \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Solver definitions used for MCSA verification and
      performance analysis.}}
  \label{tab:spn_solver_defs}
\end{table}
For the energy group structures,
Table~\ref{tab:spn_group_structure} gives the lower bounds of each
group in eV (an implicit upper bound of 20 MeV is assumed).
\begin{table}[h!]
  \begin{center}
    \begin{tabular}{cl}\hline\hline
      \multicolumn{1}{c}{\textbf{Number of Groups}} & 
      \multicolumn{1}{l}{\textbf{Lower Bounds (eV)}} \\
      1 & \{ \sn{1}{-5} \} \\
      2 & \{ \sn{1}{-1}, \sn{1}{-5} \} \\
      4 & \{ \sn{1}{1}, \sn{1}{0}, \sn{1}{-1}, \sn{1}{-5} \} \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Energy group lower bounds for the multigroup fuel
      assembly criticality problem in electron volts.}}
  \label{tab:spn_group_structure}
\end{table}

\subsection{Performance Results}
\label{subsec:spn_comparison_results}
Using the same set of solver parameters used for verification, MCSA
will be directly compared to the production Krylov solvers in terms of
both iterative performance and CPU timing. In addition to the
parameters used for the verification analysis, artificial absorption
was also introduced with a value of 0.2 for the MCSA calculations that
used to collision esitmator as CPU timing performance was improved
without a loss in iterative performance. Adding this absorption did
not change the numerical results of the
calculation. Figure~\ref{fig:spn_comparison_iterations} gives the
average number of linear solver iterations required to converge a
single eigenvalue iteration of the fuel assembly problem as a function
of the number of energy groups.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_iters.pdf}
  \end{center}
  \caption{\textbf{Average number of iterations required to converge
      each eigenvalue iteration for the fuel assembly problem as a
      function of energy groups.} \textit{All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:spn_comparison_iterations}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per iteration in seconds for the
      fuel assembly problem as a function of energy groups.}
    \textit{All linear solver iterations over all eigenvalue
      iterations were used to compute the averge. All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:spn_comparison_time}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/spn_equations/solver_p_time.pdf}
  \end{center}
  \caption{\textbf{Average CPU time per iteration in seconds for the
      fuel assembly problem as a function of energy groups with
      preconditioning time included for the MCSA methods.}
    \textit{All linear solver iterations over all eigenvalue
      iterations were used to compute the averge. All methods were
      preconditioned with identical ILUT parameters. For MCSA, several
      variations of fixed point iterations and estimators were used.}}
  \label{fig:spn_comparison_prec_time}
\end{figure}
